Kernel Development Using Rust: A Systematic Literature Review

### 2024-05-28 paper.tex
Update on Overleaf.
Updates from the holiday weekend

\begin{document}
\title{Kernel Development Using Rust: A Systematic Literature Review}
%\title{}

@@

\renewcommand{\shortauthors}{Panter et al.}

\begin{abstract}
    \textit{Context:}
    TODO:
    \textit{Objective:}
    TODO:
    \textit{Method:}
    TODO:
    \textit{Results:}
    TODO:
    \textit{Conclusions:}
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Memory safety, Rust, Kernel, Operating System}

@@

The migration of legacy C code bases to memory-safe programming languages like Rust is of paramount importance in modern software development, aiming to mitigate common security vulnerabilities and memory-related errors. This paper presents a systematic literature review (SLR) focusing on strategies and methodologies for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe languages the operating system kernel. Our study aims to provide a comprehensive overview of existing research, identify gaps, and suggest future directions in this domain. Through a rigorous search process, we synthesized relevant studies and extracted key findings to offer insights into effective approaches for ensuring memory safety when working closely with hardware.

@@

Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the inclusion and exclusion criteria based on our research questions outlined in section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published in conferences or journals, and within the time frame of 2019 and 2024. The published papers should describe using the Rust programming language for either developing a new kernel, extending an existing kernel, or authoring drivers. We will include any type of kernel architecture including monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as the paper is using the Rust programming language in some way. Papers that describe solutions that reside 100\% in user space will be excluded.

@@

Once we had our initial collection of papers from the databases we merged all duplicate papers into one record and then started our first pass which consisted of reading the title and abstract and either marking the paper as \textit{include} or \textit{exclude}. Once  the initial screening was completed we were left with 26 potential papers that we needed an in depth reading.

@@

Table~\ref{tab:RQ1} summarizes our findings regarding what existing approaches researchers are using to integrate rust into the kernel.

Boos et al.~\cite{Boos2020-zh} is about an experimental operating system that is aimed at addressing the state spill problem. State spill happens when a single service in an operating system can harbor a state change, induced by interacting with other services, that can eventually cause a system crash or system instability. The operating system is built in Rust and leverages the properties of Rust to construct a safer system. Theseus operates in a single address space and single privilege level and uses properties of the Rust programming language to realize isolation instead of relying on hardware. The design of the OS uses a novel cell based structure where ownership of memory and resources is enforced by the compiler. In short the Theseus OS is an operating system that was designed to match the Rust language instead of the more traditional route of matching the hardware.

Chen X et al.~\cite{Chen2023-wb} primary focused on developing a micro-kernel in Rust that could be formally verified by using both the liner type system or rust and a SMT solver. The new micro-kernel named Atmosphere follows the tradition of other micro-kernels by pushing most kernel functionally to users-space thus limiting the surface area that needs to be proved. The authors were able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified microkernels SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively.

Lankes et al.~\cite{Lankes2019-cm} detailed their experiences with developing a Unikernel in Rust. Originally written in C and called HermitCore the authors ported the Kernel to the Rust programming language in order to take advantage of Rust's memory safe features. The new Unikernel consists of only 3.27\% unsafe Rust with the rest of the code base consisting of safe Rust.

Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel what can run programs written in any language that can compile down to WASM.

Liang et al.~\cite{Liang2021-bo} developed a new operating system named Rustpi with a micro-kernel architecture implemented in Rust to explore how modern language features can help with building reliable systems.

\begin{table}
    \begin{tabular}{||l|l|l||}
    \hline
    Approach & Papers & Operating System in Rust\\
    \hline\hline
    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg}, \cite{Oikawa2023-ms} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe}, \cite{Narayanan2020-gs}, \cite{Narayanan2019-fd} & Atmosphere, Redox, Redleaf\\
    Embedded & \cite{Culic2022-bk}, \cite{Vishnunaryan2022-yd} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern}, HarSaRK \\
    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}, \cite{Sung2020-bb}  & RustyHermit, Theseus \\
    Exokernel & \cite{Li2024-yb} & W-Kernel \\
    \hline
  \end{tabular}
  \caption{Approaches and Methodologies for Rust in the Kernel}
    \label{tab:RQ1}
\end{table}

\subsection{RQ2  What are the performance implications of using Rust for operating system
      kernel development in terms of throughput, latency, and resource utilization?}

Comparing performance, throughput, latency, and resource utilization between different systems is very difficult. For example several of the operating systems written in Rust such as Tock are a complete rethinking of how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had the benefit of decades of incremental performance improvements like components that are written in C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are not directly meaningful due to the vastly different goals of each system. Therefore, we will focus more on the overall efforts that are specific to the Rust programming language and the challenges presented. We summarize our findings in table~\ref{tab:RQ2}.

@@

\begin{table}
    \begin{tabular}{||l|l|l||}
    \hline
    No. & Implication & Studies that Reported the challenge\\
    \hline\hline
    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
    2 & Throughput & \cite{Gonzalez2023-ek}\\
    3 & Latency & \cite{Culic2022-bk} \\
    4 & Resource Utilization & 0 \\
    \hline
  \end{tabular}
  \caption{Performance Implications of using Rust in the Kernel}
    \label{tab:RQ2}
\end{table}

@@

            \item Support typed assembly language for Rust
            \item Support trusted build environments 
            \item Provide software-only stack guard with extensible probing interface
            \item Develop zero-copy serialization of “plain- old” data structures
        \end{itemize}
        \\
        \hline
        3 & Soundness~\cite{Klimt2023-ob} &
        \begin{itemize}
            \item Unsynchronized Global State - any use of mutable statics is unsafe
            \item C-Style Abstractions - Use rust style abstractions to properly encapsulate internal unsafe usage of raw pointers
            \item Aliasing Mutable References - Giving out raw pointers to memory that is also referenced mutably
            \item Re-implementing Memory Access - instead of accessing specific memory regions in assembly, creating and using references to the whole region is preferable
        \end{itemize}
        \\
        \hline
        4 & Panics \cite{Ma2023-ef},\cite{Burtsev2021-mh} &
        \begin{itemize}
            \item Support extendable, no\_std unwind library
            \item Stack unwinding in embedded environments
        \end{itemize}
        \\
        \hline
        5 & C Interop \cite{Miller2019-xm} &
        \begin{itemize}
            \item Kernel interfaces, while designed for extensibility, are not designed for type safety
            \item Hybrid Code Flow. The rust compiler can not track ownership when switching between modules written in C and Rust
        \end{itemize}
        \\
    \hline
  \end{tabular}
  \caption{Challenges Unique to the Rust Programming Language}
    \label{tab:RQ3}
\end{table*}

Ayers et al.~\cite{Ayers2022-sf} focused on the size of rust binaries in the embedded environment. Their work primary focused on the size of binaries produced using the embedded OS Tock. In addition to the identified causes of binary growth the authors have the following 5 recommendations when using Rust in a size constrained environment:
\begin{itemize}
  \item Minimize Length + Instantiations of Generic Code
  \item Use Trait Objects Sparingly
  \item Don't Panic
  \item Carefully Use Compiler Generated Support Code
  \item Don't use static mut
\end{itemize}

Burtsev et al.~\cite{Burtsev2021-mh} explores what is missing in the Rust programming language to help solve the isolation problem in an operating system. Currently Rust lacks the ability to express isolation in the heap without external support. The RedLeaf experiential operating system relied on an complex interface definition language (IDL) to enforce isolated heaps. The paper enumerates several properties of the Rust Language that could help with isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}. The authors argue that with these changes or inclusions to the Rust programming language developing operating system kernels would be much easier and safer.

Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing Theseus, an operating system written in Rust. First they found that it is impossible to write a complete operating system in 100\% safe rust. For example when writing a memory management system raw pointers must be used in order to modify the hardware. Despite the limitations imposed by the hardware the authors describe how Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness. By leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be achieved at a higher level than what could be done in C. The authors also detailed some of the limitations of intralingual design such as not being as expressive as many other formal verification techniques due to the limited invariant that can be enforced by the type system. One of the most important lessons learned was the insight that a linear type system itself cannot guarantee uniqueness of the resource represented, such when a memory resources may overlap. The authors introduce the idea of using a hybrid approach of verification where they leverage both the linear type system and an SMT solver. The authors also explore the bootstrapping problem with rust systems. How do we ensure that ownership of memory is sound in the operating system itself? In operating systems written in C the kernel provides the ownership root to applications running on top, finding a new ownership root is a open research question with regards to using Rust for operating system development.

@@

Summary of key findings and insights obtained from the systematic literature review.
Concluding remarks on the significance of memory-safe programming languages in mitigating security vulnerabilities.
Closing thoughts on the importance of minimizing unsafe blocks during code migration.

\section{Threats to Validity}
\balance

>>>>

\begin{document}
\title{Kernel Development Using Rust: A Systematic Literature Review}

\author{Anonymous Authors}

@@

%\renewcommand{\shortauthors}{Panter et al.}

\begin{abstract}

  \textit{Context:} The integration of Rust into kernel development is a transformative endeavor aimed
  at enhancing system security and reliability by leveraging Rust's strong memory safety
  guarantees. This systematic literature review examines the current state of research and practice
  regarding the use of Rust in the kernel development.
  \textit{Objective:} We aim to find the current edge of research regarding using Rust in Kernel
  development so we can push the research forward and reduce the number of memory safety
  vulnerabilities in one of the most critical pieces of software that underpins all modern
  applications.
  \textit{Method:} By analyzing a broad spectrum of studies, we identify the advantages Rust offers,
  such as reducing vulnerabilities like buffer overflows and use-after-free errors, which are
  prevalent in C-based kernel development. The review also highlights the challenges faced,
  including integration hurdles with existing C codebases, performance considerations, and the need
  for community consensus on Rust's adoption.
  \textit{Results:} Our findings suggest that while the initial implementations of Rust in the kernel
  shows promising results in terms of safety and stability, significant challenges remain. These
  include achieving seamless interoperability with existing kernel components, maintaining
  performance, and ensuring adequate support and tooling for developers.
  \textit{Conclusions:} In conclusion, this review provides a comprehensive overview of the
  potential and challenges of incorporating Rust into kernel development. It underscores the need for
  continued research and practical implementation efforts to fully realize the benefits of Rust. By
  addressing these challenges, the integration of Rust could mark a significant step forward in the
  evolution of operating system development towards safer and more reliable systems.

\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Memory safety, Rust, Kernel, Operating System, Linux}

@@

Most operating system kernels are predominantly written in the C programming language with bits of
assembly. C has been favored for its low-level capabilities and performance efficiency, crucial for
kernel development. However, the inherent lack of memory safety in C has led to numerous security
vulnerabilities, including buffer overflows and use-after-free errors, which have plagued operation
system development for decades~\cite{noauthor_undated-bf}. Addressing these vulnerabilities is
paramount to enhancing the security and reliability of all operating systems.

Rust, a systems programming language developed by Mozilla, has garnered significant attention for
its strong emphasis on memory safety without sacrificing performance. Rust's ownership model and its
compile-time checks effectively prevent common programming errors that lead to security
vulnerabilities. What makes Rust unique is the fact that it accomplishes these tasks without the use
of a garbage collector, making it an ideal candidate for kernel development where safety and
performance are critical.

The potential integration of Rust into kernel development represents a significant shift in the
landscape of operating system development. This shift prompts a comprehensive evaluation of both the
opportunities and challenges associated with using Rust in such a foundational component of modern
computing. This systematic literature review aims to synthesize existing research and practical
experiences related to the use of Rust in the kernel, providing a detailed understanding of
the current state of this emerging field.

This paper presents a systematic literature review (SLR) focusing on strategies and methodologies
for integrating Rust into one of the most fundamental areas that are typically dominated by unsafe
languages the operating system kernel. Our study aims to provide a comprehensive overview of
existing research, identify gaps, and suggest future directions in this domain. Through a rigorous
search process, we synthesized relevant studies and extracted key findings to offer insights into
effective approaches for ensuring memory safety when working closely with hardware.

@@

Following the guidelines outlined by Kitchenham and Charters~\cite{Stuart2007-cc} we set the
inclusion and exclusion criteria based on our research questions outlined in
section~\ref{sec:researchQuestions}. We only consider papers that are written in English, published
in conferences or journals, and within the time frame of 2019 and 2024. The published papers should
describe using the Rust programming language for either developing a new kernel, extending an
existing kernel, or authoring drivers. We will include any type of kernel architecture including
monolithic kernel, microkernel, or unikernel in both the embedded and non-embedded space as long as
the paper is using the Rust programming language in some way. Papers that describe solutions that
reside 100\% in user space will be excluded. Early efforts to use Rust in kernel development
were impacted by changes and updates in both the Rust language and compiler and relied on unstable
features and thus are excluded from this review.

@@

Once we had our initial collection of papers from the databases we merged all duplicate papers into
one record and then started our first pass which consisted of reading the title and abstract and
either marking the paper as \textit{include} or \textit{exclude}. Once the initial screening was
completed we were left with 26 potential papers that we needed an in depth reading with an
additional 2 papers found during forward and backward snowballing.

@@

The integration of Rust into operating system kernel development has garnered significant attention
due to Rust's promise of memory safety and concurrency guarantees. Several approaches and
methodologies have emerged to leverage Rust for kernel development, each addressing different
aspects of kernel functionality and integration.  We found several approaches that researches are
taking to integrating Rust into kernel development.  The first approach is a greenfield approach
were the operating system architecture is redesigned and built from the ground up in order to fully
leverage the Rust programming language. The second approach is the incremental integration of Rust
into existing kernels, such as the Linux
kernel~\cite{The_kernel_development_community_undated-iw}. This methodology involves gradually
rewriting or augmenting specific components of the kernel with Rust while maintaining compatibility
with the existing C codebase. Table~\ref{tab:RQ1} summarizes our findings regarding what existing
approaches researchers are using to integrate rust into the kernel.

For the first approach, Boos et al.~\cite{Boos2020-zh} created an experimental operating system
named Theseus that operates in a single address space and single privilege level and uses properties
of the Rust programming language to realize isolation instead of relying on hardware. This novel
approach pushes some of the roles that a traditional operating system would take onto the
compiler. The design of Theseus uses a novel cell based structure where ownership of memory and
resources are enforced by the compiler, thus avoiding these checks during runtime to get improved
performance. In addition to memory safety, Theseus has a goal of addressing the state spill
problem. State spill happens when a single service in an operating system can harbor a state change,
induced by interacting with other services. That state change can eventually cause a system crash or
system instability at some point in the future making it very difficult to track down the problem
due to the separation between when the problem occurred and when it was observed. Theseus OS is an
operating system that was designed to match the Rust language instead of the more traditional route
of matching the hardware.

While single address space operating systems present a unique opportunity to push more
responsibility into the compiler a more traditional micro-kernel implementation was taken by Chen X
et al.~\cite{Chen2023-wb} and Liang et al.~\cite{Liang2021-bo}. By pushing most kernel functionally
to user space the authors limited the surface area that needs to be proved. Chen X et al. took
their implementation one step further by formally verifying their new micro-kernel named Atmosphere
by using both the liner type system of Rust in combination with a SMT solver. Chen X et al. were
able to get a 7.5:1 proof-to-code ratio which is higher that other formerly verified micro-kernels
SeL4 and CeriKOS, which both have proof to code ratio of 19:1 and 20:1 respectively. While this
approach was novel drivers running in userspace are still not formally verified and don't have to be
written in Rust.

Unikernels present an ideal surface for prototyping Rust. Unikernels are designed to do one thing
and one thing only, thus the limiting the complexity that a general purpose operating system has to
deal with. This reduced complexity was explored by Lankes et al.~\cite{Lankes2019-cm}. The authors
detailed their experiences with rewriting an existing Unikernel, HermitCore, in Rust. The new
Unikernel named RustyHermit consists of only 3.27\% unsafe Rust with the rest
of the code base consisting of safe Rust dramatically decreasing the surface area where memory errors
could originate from.

Finally, Li and Sato~\cite{Li2024-yb} explored using Rust to write an exokernel named W-Kernel. The authors
proposed a novel architecture for an OS by embedding a WebAssembly(WASM) runtime into the kernel
what can run programs written in any language that can compile down to WASM.

\begin{table*}
    \begin{tabular}{||l|l|l||}
    \hline
    Approach & Papers & Operating System in Rust\\
    \hline\hline
    Monolithic  & \cite{The_kernel_development_community_undated-iw}, \cite{Li2019-ru}, \cite{Miller2021-pg}, \cite{Oikawa2023-ms} & \href{https://docs.kernel.org/rust/}{Linux kernel v6.1+}\\
    Micro-kernel & \cite{Chen2023-wb},\cite{Liang2021-bo}, \cite{Liu2024-xe}, \cite{Narayanan2020-gs}, \cite{Narayanan2019-fd} & Atmosphere, Redox, Redleaf\\
    Embedded & \cite{Culic2022-bk}, \cite{Vishnunaryan2022-yd} & \href{https://github.com/tock/tock}{Tock}, \href{https://hubris.oxide.computer/}{Hubris}, \href{https://www.drone-os.com/}{Drone}, \href{https://bern-rtos.org/}{Bern}, HarSaRK \\
    Unikernel & \cite{Lankes2019-cm},  \cite{Boos2020-zh}, \cite{Ijaz2023-da}, \cite{Sung2020-bb}  & RustyHermit, Theseus \\
    Exokernel & \cite{Li2024-yb} & W-Kernel \\
    \hline
  \end{tabular}
  \caption{Approaches and Methodologies for Rust in the Kernel}
    \label{tab:RQ1}
\end{table*}

\subsection{RQ2  What are the performance implications of using Rust for operating system
      kernel development in terms of throughput, and  latency?}

Comparing performance, throughput and latency between different systems is very difficult. For
example several of the operating systems written in Rust such as Tock are a complete rethinking of
how an operating system is designed and Rust for Linux (RFL) is still relatively new and has not had
the benefit of decades of incremental performance improvements like components that are written in
C. Additionally, direct comparisons between a Monolithic kernel, Micro-kernel, and Unikernel, or
comparisons between a Real Time Operation System (RTOS) and a General Purpose operating System are
not directly meaningful due to the vastly different goals of each system. Therefore, we will focus
more on the overall efforts that are specific to the Rust programming language and the challenges
presented. We summarize our findings in table~\ref{tab:RQ2}.

@@

\begin{table}
    \begin{tabular}{||l|l|l||}
    \hline
    No. & Implication & Studies that Reported the challenge\\
    \hline\hline
    1 & Performance & \cite{Gonzalez2023-ek}, \cite{Li2024-be}, \cite{Ma2023-ef}\\
    2 & Throughput & \cite{Gonzalez2023-ek}\\
    3 & Latency & \cite{Culic2022-bk} \\
    \hline
  \end{tabular}
  \caption{Performance Implications of using Rust in the Kernel}
    \label{tab:RQ2}
\end{table}

@@

            \item Support typed assembly language for Rust
            \item Support trusted build environments
            \item Provide software-only stack guard with extensible probing interface
            \item Develop zero-copy serialization of “plain- old” data structures
        \end{itemize}
        \\
        \hline
        3 & Soundness~\cite{Klimt2023-ob} &
        \begin{itemize}
            \item Unsynchronized Global State - any use of mutable statics is unsafe
            \item C-Style Abstractions - Use rust style abstractions to properly encapsulate internal unsafe usage of raw pointers
            \item Aliasing Mutable References - Giving out raw pointers to memory that is also referenced mutably
            \item Re-implementing Memory Access - instead of accessing specific memory regions in assembly, creating and using references to the whole region is preferable
        \end{itemize}
        \\
        \hline
        4 & Panics \cite{Ma2023-ef},\cite{Burtsev2021-mh} &
        \begin{itemize}
            \item Support extendable, no\_std unwind library
            \item Stack unwinding in embedded environments
        \end{itemize}
        \\
        \hline
        5 & C Interop \cite{Miller2019-xm}, \cite{Li2021-xo} &
        \begin{itemize}
            \item Kernel interfaces, while designed for extensibility, are not designed for type safety
            \item Hybrid Code Flow. The rust compiler can not track ownership when switching between modules written in C and Rust
        \end{itemize}
        \\
    \hline
  \end{tabular}
  \caption{Challenges Unique to the Rust Programming Language}
    \label{tab:RQ3}
\end{table*}

The development of operating system kernels in Rust introduces several unique challenges and
limitations compared to more traditional languages like C. These challenges arise from Rust’s strict
safety guarantees, its relatively recent adoption in systems programming, and the inherent
complexities of kernel development. This section discusses these challenges, limitations, and the
lessons learned from various projects and research efforts.

A primary concern for programming languages that are used for operating system development are the
size of the binary especially in an embedded environment.  Ayers et al.~\cite{Ayers2022-sf} focused
on reducing size of binaries produced by the rust compiler while working with Tock OS. They were
able to identify several  causes of binary growth that are specific to the Rust programming language
and have the following 5 recommendations when using Rust in a size constrained environment:
\begin{itemize}
  \item Minimize Length + Instantiations of Generic Code
  \item Use Trait Objects Sparingly
  \item Don't Panic
  \item Carefully Use Compiler Generated Support Code
  \item Don't use static mut
\end{itemize}

Burtsev et al.~\cite{Burtsev2021-mh} explored what is missing in the Rust programming language to
help solve the isolation problem in an operating system. Currently Rust lacks the ability to express
isolation in the heap without external support. For example, the RedLeaf experiential operating
system which is written in Rust relied on an complex interface definition language (IDL) to enforce
isolated heaps. The paper enumerates several properties of the Rust Language that could help with
isolation with regards to operating system development that are detailed in table~\ref{tab:RQ3}. The
authors argue that with these changes or inclusions to the Rust programming language developing
operating system kernels would be much easier and safer.

Klimt et al.~\cite{Klimt2023-ob} details the lessons learned and challenges when implementing
Theseus. First they found that it is impossible to write a complete operating system in 100\% safe
Rust. For example when writing a memory management system raw pointers must be used in order to
modify the hardware. Despite the limitations imposed by the hardware the authors describe how
Theseus leveraged intralingual design to maximize the compilers role in enforcing correctness. By
leveraging Rusts type system and borrow checker memory safety and correct ownership transfer can be
achieved at a higher level than what could be done in C. The authors also detailed some of the
limitations of intralingual design such as not being as expressive as many other formal verification
techniques due to the limited invariant that can be enforced by the type system. One of the most
important lessons learned was the insight that a linear type system itself cannot guarantee
uniqueness of the resource represented, such when a memory resources may overlap. The authors
introduce the idea of using a hybrid approach of verification where they leverage both the linear
type system and an SMT solver. The authors also explore the bootstrapping problem with rust
systems. How do we ensure that ownership of memory is sound in the operating system itself? In
operating systems written in C the kernel provides the ownership root to applications running on
top, finding a new ownership root is a open research question with regards to using Rust for
operating system development.

\section{Threats to Validity}

In conducting this systematic literature review on the use of Rust for operating system development,
several potential threats to validity must be acknowledged. These threats are categorized into
internal, external, and conclusion validity, each presenting distinct challenges to the robustness
and reliability of our findings.

Definition Ambiguity: The term "Kernel development in Rust" encompasses a wide range of activities,
from minor module development to complete subsystem rewrites. Variations in how researchers and
developers interpret this integration could lead to inconsistent findings. Selection Criteria: The
inclusion and exclusion criteria for selecting relevant studies might inadvertently bias the review
towards certain types of research, such as those reporting successful integration while
underrepresenting studies detailing challenges and failures. Compared to C Rust is a brand new
language that is still finding its way in the world.

Internal Validity: Internal validity refers to the degree to which the review accurately establishes
causal relationships. Potential threats include:

\begin{itemize}
  \item Publication Bias: Studies demonstrating positive outcomes of using Rust for kernel
    development may be more likely to be published and cited, leading to an overestimation of its
    benefits.
  \item Quality of Studies: The methodological rigor of the included studies varies, and some may
    suffer from design flaws or biases that were not adequately controlled for, thus affecting the
    reliability of their findings.
\end{itemize}

External Validity: External validity pertains to the generalizability of the review's findings
beyond the specific context of the reviewed studies. Threats to external validity include:

\begin{itemize}
  \item Context-Specific Findings: Many studies focus on specific kernel modules or use-cases, which
    may not be representative of the broader kernel environment. The success of Rust in isolated
    components does not necessarily translate to the entire kernel.
  \item Temporal Changes: The rapidly evolving nature of both the Rust programming language and the
    Linux kernel means that findings from older studies may no longer be applicable, as improvements
    and changes in both domains can alter the landscape significantly.
\end{itemize}

Conclusion Validity: Conclusion validity involves the degree to which the conclusions drawn from the
review are credible and supported by the data. Potential threats include:

\begin{itemize}
  \item Heterogeneity of Studies: The diverse methodologies, metrics, and contexts of the included studies
    can lead to challenges in synthesizing findings and drawing unified conclusions. For example, it
    is very difficult to directly compare a micro-kernel, Unikernel, and hybrid Kernel.
  \item Reviewer Bias: Personal biases of the reviewers in interpreting data and making judgments
    about study quality and relevance may skew the results
\end{itemize}

Efforts were made to minimize this through predefined criteria and multiple reviewers, but some
degree of subjectivity is inevitable.  In addressing these threats, we employed rigorous methods for
study selection, data extraction, and analysis, and we remain transparent about the
limitations. Future reviews could mitigate these threats further by incorporating a broader range of
sources, using more precise definitions, and employing meta-analytic techniques to quantitatively
synthesize findings. Despite these threats, our review provides a valuable synthesis of the current
state of research on the use of Rust in kernel development, highlighting both its potential benefits
and the challenges that need to be addressed.

@@

The integration of Rust into  kernel space represents a significant evolution in operating system
development, promising to address long-standing issues related to memory safety and system
reliability. This systematic literature review has synthesized the current state of research and
practice concerning the use of Rust in multiple types of kernels including the venerable Linux
kernel, highlighting both the progress made and the challenges that remain.

Our review indicates that Rust's strong guarantees of memory safety, enabled by its ownership model
and strict compile-time checks, offer a compelling advantage over traditional C-based kernel
development. These features have the potential to reduce common vulnerabilities such as buffer
overflows and use-after-free errors, which are prevalent in C and have historically led to critical
security exploits. Several case studies and prototype implementations have demonstrated that Rust
can be successfully integrated into the kernel, providing safer interfaces and reducing the
incidence of memory-related bugs without incurring significant performance penalties.

However, the adoption of Rust for kernel development is not without its challenges. Truly the
biggest challenge is existing codebases and mountains of legacy code. With extensive use of C and
reliance on specific C idioms and low-level programming techniques researchers have substantial
integration hurdles to overcome. Any efforts to rewrite substantial portions of an existing kernel
such as Linux in Rust are constrained by the need for interoperability with existing C code and the
necessity to maintain the kernel's performance characteristics. Furthermore, the operating system
development community must reach consensus on Rust's role and ensure that sufficient tooling,
documentation, and support are available for developers.

In conclusion, while Rust's incorporation into the kernel space is still in its nascent stages, the
initial results are promising. The potential for enhanced security and stability aligns with the
long-term goals of kernel development, and continued research, coupled with practical implementation
efforts, will be crucial in realizing these benefits. Future work should focus on addressing
integration challenges, refining interoperability mechanisms, and expanding the body of empirical
evidence on Rust's impact within the kernel environment. As the community navigates these
challenges, the evolution of Rust in the kernel may well mark a transformative period in the pursuit
of safer, more reliable operating systems.

\balance
